Coalescent inference of mutation spectrum histories from sample frequency spectra

### 2019-09-10 manuscript/main.tex
promote vectors to spectrum matrices in TeX

\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\lngamma}{\ln\!\Gamma}

@@

Given $n$ sampled haplotypes and a demographic history $\eta(t)$ (with time measured retrospectively from the present), we show in Appendix \ref{sec:appendix:xi} that the expected SFS $\boldsymbol \xi = (\xi_1, \xi_2,\dots, \xi_{n-1})$ is a linear transform of the mutation intensity history $\mu(t)$:
\begin{equation}
  \label{eqn:transform}
\boldsymbol \xi = \mathcal{L}_{n,\eta}\mu,
\end{equation}
where $\mathcal{L}_{n,\eta}$ is a finite-rank linear operator that maps infinite-dimensional mutation intensity histories to $(n-1)$-dimensional SFS vectors.
It is dependent on $n$ and $\eta$, which we take to be fixed.

@@

For numerical implementation we consider finite-dimensional approximations of $\eta(t)$ and $\mu(t)$ as piecewise constant on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$.
We take the epoch boundaries as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
Let $\boldsymbol y = (y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $\boldsymbol z = (z_1,\dots,z_m)$ denote the constant mutation intensity $\mu(t)$ during each epoch.
In Appendix \ref{sec:appendix:pcsws} we show the linear transform \eqref{eqn:transform} reduces to the matrix equation
\begin{equation}
\label{eqn:transform_discrete}
\boldsymbol \xi = L_{n, \boldsymbol y} \boldsymbol z,
\end{equation}
where the $(n-1)\times m$ matrix $L_{n, \boldsymbol y}$ is fixed given a fixed demographic history $\boldsymbol y$.

In linkage equilibrium the log likelihood function of the mutation intensity history $\boldsymbol z$ given an observed SFS vector $\boldsymbol x = (x_1,\dots,x_{n-1})$ and demographic history $\boldsymbol y$ is given by the Poisson random field approximation \citep{}
\[
\ell(\boldsymbol z; \boldsymbol x, \boldsymbol y, n) = \boldsymbol x^\intercal\log(L_{n, \boldsymbol y} \boldsymbol z) - \boldsymbol 1^\intercal L_{n, \boldsymbol y} \boldsymbol z,
\]
where $\log(\cdot)$ operates elementwise, and we've dropped a constant term wrt $\boldsymbol z$.

%WD more inverse problem intro
The inverse problem \eqref{eqn:transform} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
In order to recover smooth histories,
we penalize the $L^p$ norms
of the derivative of $\mu$
\[
\left\| \frac{d \mu}{d t} \right\|_p^p
= \int_0^\infty\left|\frac{d\mu(t)}{dt}\right|^p dt,
\]
for $p=1$ and 2.
With time discretized \eqref{eqn:transform_discrete},
this becomes
\[
\left\|\Delta\boldsymbol z \right\|_p^p.
\]
where $\Delta$ denotes the first difference matrix.
Our cost function (penalized log-likelihood) is then
\begin{equation}
\label{eqn:penalized}
C(\bm z)
%\tilde\ell(\boldsymbol z)
= -\ell(\boldsymbol z) + \lambda \left(\alpha\left\|\Delta\boldsymbol z\right\|_1 + \frac{(1-\alpha)}{2}\left\|\Delta\boldsymbol z\right\|_2^2\right),
\end{equation}
where $\lambda$ tunes the regularization strength, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
\textcolor{blue}{Kam note: probably we need a ``Ridge'' (Tikhonov)
term like the following to deal with small singular values:}
\[
\lambda' \| \bm z \|_2^2
\]

>>>>

\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\lngamma}{\ln\!\Gamma}
\DeclareMathOperator{\Tr}{Tr}

@@

Given $n$ sampled haplotypes and a demographic history $\eta(t)$ (with time measured retrospectively from the present), we show in Appendix \ref{sec:appendix:xi} that the expected SFS $\boldsymbol \xi = (\xi_1, \xi_2,\dots, \xi_{n-1})$ is a linear transform of the mutation intensity history $\mu(t)$:
\begin{equation}
  \label{eqn:transform}
\boldsymbol \xi = \mathcal{L}_{n,\eta}\mu(t),
\end{equation}
where $\mathcal{L}_{n,\eta}$ is a finite-rank linear operator that maps infinite-dimensional mutation intensity histories to $(n-1)$-dimensional SFS vectors.
It is dependent on $n$ and $\eta$, which we take to be fixed.

@@

In previous work \citep{Kelley, cancer, etc}, mutation spectra have been parameterized by grouping variants into mutation types according to local $k$-mer nucleotide context, and the number of variants called in each mutation type is taken as a proxy integrating an individual's ancestral mutational input in each type.
With $k$ odd and the variant registered in the central position there are $\kappa_k=2\times3\times4^{k-1}$ mutation types after collapsing by strand symmetry.
For example, there are $96$ triplet mutation types ($k=3$).
Here, we use the same $k$-mer mutation type parameterization to promote the $(n-1)$-element expected SFS vector $\xi$ to the $(n-1)\times\kappa_k$ expected $k$-SFS matrix $\Xi$.
Similarly, the mutation intensity history function $\mu(t)$ is promoted to the $\kappa_k$-element mutation spectrum history $\boldsymbol\mu(t)$, with each element giving the mutation intensity history function for one mutation type.
Equation \eqref{eqn:transform} becomes
\begin{equation}
  \label{eqn:transform_matrix}
\Xi = \mathcal{L}_{n,\eta}\boldsymbol\mu^\intercal(t),
\end{equation}

For numerical implementation we consider finite-dimensional approximations of $\eta(t)$ and $\boldsymbol\mu(t)$ as piecewise constant functions of time on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$.
We take the epoch boundaries as fixed parameters, and in practice make them dense so as to approximate infinite-dimensional histories.
Let $\boldsymbol y = (y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let the $m\times\kappa_k$ matrix $Z$ denote the constant mutation intensity during each epoch (rows) for each mutation type (columns).
In Appendix \ref{sec:appendix:pcsws} we show the linear transform \eqref{eqn:transform_matrix} reduces to the matrix equation
\begin{equation}
\label{eqn:transform_discrete}
\Xi = L_{n, \boldsymbol y} Z,
\end{equation}
where the $(n-1)\times m$ matrix $L_{n, \boldsymbol y}$ is fixed given a fixed demographic history $\boldsymbol y$.

In linkage equilibrium the log likelihood function of the mutation spectrum history $Z$ given an observed $k$-SFS matrix $X$ and demographic history $\boldsymbol y$ is given by the Poisson random field approximation \citep{}
\[
\ell(Z; X, \boldsymbol y, n) = \Tr(X^\intercal\log(L_{n, \boldsymbol y} Z)) - \boldsymbol 1^\intercal L_{n, \boldsymbol y} Z\boldsymbol 1,
\]
where $\log(\cdot)$ operates elementwise, and we've dropped a constant term wrt $Z$.

%WD more inverse problem intro
The inverse problem \eqref{eqn:transform_discrete} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
In order to recover smooth histories,
we penalize the $L^p$ norms
of the time derivative of $\boldsymbol\mu(t)$
\[
\sum_{i=1}^{\kappa_k}\left\| \frac{d \mu_i(t)}{d t} \right\|_p^p
= \sum_{i=1}^{\kappa_k}\int_0^\infty\left|\frac{d\mu_i(t)}{dt}\right|^p dt,
\]
for $p=1,2$.
With piecewise constant histories as in \eqref{eqn:transform_discrete} this becomes
\[
\left\|\Delta Z \right\|_p^p.
\]
where $\Delta$ denotes the first difference matrix.
Our cost function (penalized log-likelihood) is then
\begin{equation}
\label{eqn:penalized}
C(Z)
%\tilde\ell(\boldsymbol z)
= -\ell(Z) + \lambda \left(\alpha\left\|\Delta Z\right\|_1 + \frac{(1-\alpha)}{2}\left\|\Delta Z\right\|_2^2\right),
\end{equation}
where $\lambda$ tunes the regularization strength, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
\textcolor{blue}{Kam note: probably we need a ``Ridge'' (Tikhonov)
term like the following to deal with small singular values:}
\[
\lambda' \| Z \|_2^2
\]
